---
title: "Preprocessing script for Nestorowa 2017"
author: "Aditya Pratapa"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
---
Load necesdsary libraries
```{r warning=FALSE,message=FALSE}
library(destiny)
library(slingshot)
library(plotly)
library(gam)
library(RColorBrewer)
```
Read input files. Both the input .txt files are from http://blood.stemcells.cam.ac.uk/data/

```{r warning=FALSE}
Nestorowa2017 <- read.table('normalisedCountsVariableGenes.txt', header = T, row.names = 1)
NestorowaCellTypes <- read.table('all_cell_types.txt', header = T, row.names = 1)
dim(Nestorowa2017)
head(NestorowaCellTypes)
```
Compute diffusion map projection of the cells
```{r warning=FALSE, fig.height=10}
logExpression <- log2(Nestorowa2017 + 1)
dm <- DiffusionMap(t(logExpression), distance = "cosine", sigma = .16)
rd2 <- cbind(DC1 = dm$DC1, DC2 = dm$DC2, DC3 = dm$DC3)

plot_ly(as.data.frame(rd2), type='scatter3d', x= ~DC1, y= ~DC2, z= ~DC3, size = 5, color = "grey")
```

Map the cell types identified using narrow gating in the original publication to the lower dimenstional representation.
```{r warning=False}
# Process the cell type file, use only the following columns instead of all the broad categories
cellTypes <- NestorowaCellTypes[,c('ESLAM','LTHSC_broad','LMPP_broad','MPP_broad','CMP_broad','MEP_broad','GMP_broad','MPP1_broad','MPP2_broad','MPP3_broad','STHSC_broad')]
# assign a cluster label to each cell type
cellTypeDict <- DataFrame()
for(cellName in colnames(Nestorowa2017)){
 cTypeName <- colnames(cellTypes)[which(cellTypes[cellName[1],] == 1)][1]
 if(is.na(cTypeName)){
   cTypeName <- 'Unknown'
}
  cellTypeDict[cellName,'Type'] <- cTypeName
}
plot_ly(as.data.frame(rd2), type='scatter3d', x= ~DC1, y= ~DC2, z= ~DC3, size = 5, color = cellTypeDict$Type, colors = 'Set2')

```
Run slingshot
```{r}
slingshotPT <- slingshot(rd2, reducedDim = rd2,clusterLabels =  cellTypeDict$Type,  start.clus = 'ESLAM', end.clus = c('MEP_broad','GMP_broad','LMPP_broad'))
```
```{r}
ssPT <- slingPseudotime(slingshotPT)
ssPT <- as.data.frame(ssPT)
plot_ly(as.data.frame(rd2), type='scatter3d', x= ~DC1, y= ~DC2, z= ~DC3, size = 5, color = ssPT$curve1)
plot_ly(as.data.frame(rd2), type='scatter3d', x= ~DC1, y= ~DC2, z= ~DC3, size = 5, color = ssPT$curve2)
plot_ly(as.data.frame(rd2), type='scatter3d', x= ~DC1, y= ~DC2, z= ~DC3, size = 5, color = ssPT$curve3)

```
According to the paper the three lineages above are as follows: curve1 is Granulocyte-Macrophage (GM), curve2 is Erythroid (E), and curve3 is Lymphoid (L). 

Find variable genes along the GM trajectory

```{r}
lineages <- c('GM','E','L')
count <- 1
for (col in colnames(ssPT)){
t <- ssPT[,col]
Y <- logExpression
var1K <- names(sort(apply(Y,1,var),decreasing = TRUE))
Y <- Y[var1K,]
# fit a GAM with a loess term for pseudotime
gam.pval <- apply(Y,1,function(z){
    d <- data.frame(z=z, t=t)
    suppressWarnings({
      tmp <- gam(z ~ lo(t), data=d)
    })
    p <- summary(tmp)[4][[1]][1,5]
    p
})

topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:10]
heatdata <- logExpression[topgenes, order(t, na.last = NA)]
heatclus <- as.factor(cellTypeDict[order(t, na.last = NA),])
heatmap(as.matrix(heatdata), Colv = NA,
        ColSideColors = brewer.pal(9,"Set1")[heatclus], labCol = FALSE)


exprData <- logExpression[,which(!is.na(t))]
colnames(exprData) <- gsub(pattern = "[.]", replacement = '_', colnames(exprData))
ptData <- data.frame(ssPT[which(!is.na(t)), col])
rownames(ptData) <- colnames(exprData)
colnames(ptData) <- 'PseudoTime'

geneData <- data.frame(sort(gam.pval, decreasing = FALSE))
colnames(geneData) <- 'VGAMpValue'
geneData[,'Variance'] <- apply(exprData[rownames(geneData),],1,var)


write.csv(x = exprData, file = paste0('ExpressionData-',lineages[count],'.csv'), quote = FALSE)
write.csv(x = ptData, file = paste0('PseudoTime-',lineages[count],'.csv'), quote = FALSE)
write.csv(x = geneData, file = paste0('GeneOrdering-',lineages[count],'.csv'), quote = FALSE)
count <- count +1
}

```

Sanity check
```{r}
rdDF <- as.data.frame(rd2)
rdDF <- rdDF[which(!is.na(ssPT$curve3)),]


plot_ly(rdDF, type='scatter3d',  x= ~DC1, y= ~DC2,  z= ~DC3,color = ptData$PseudoTime)

plot_ly(as.data.frame(rd2), type='scatter3d', x= ~DC1, y= ~DC2, z= ~DC3, size = 5, color = ssPT$curve3)


```